Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2015-11-13 docs/ms_sageAbundance.Rmd
edits to Andy's abstract edits
updates refs for Andy's edits

Abstract
--------
New remote sensing technology is producing ecological data at unprecedented spatial scales. 
These massive data offer an opportunity to study ecological processes at the landscape and regional scales that have traditionally been observed only at the level of small study plots. 
However, harnessing the power of new spatial data requires statistical models that can both rigorously account for spatial autocorrelation and can also be fit within a reasonable amount of computation time. 
Here we demonstrate how recent advances in the modeling of animal populations can be applied to newly available remotely-sensed species-specific estimates of plant cover at the landscape scale.  
Using modern, computationally efficient model reparameterizations we fit a spatiotemporal population model to a 27 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming with rigorous accounting of spatial autocorrelation.
We include interannual variation in precipitation and temperature as covariates in the model to investigtate how climate affects the cover of this species.
We then use the model to forecast the future abudance of sagebrush at the landscape scale under projected climate change. 
Our analysis provides spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale. 
Critically our large-scale and long-term predictions are rooted in small-scale and short-term population dynamics and so serve as a useful alternative to the predictions offered by most species distribution models that do not include population dynamics at all. 
We expect the approach we develope here will become an indispensable method for using new remote sensing data to model how populations respond to environmental changes, especially climate change, that play out at spatial scales far greater than the average study plot.

@@

Introduction
-------------
Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology [@Clark2001; @Petchey2015].
Population models are ideally suited for meeting such a challenge because they allow inference on population abundance and dynamics (year-to-year changes in abundance) in relation to climate drivers [@Hare2010; @Adler2012; @Ross2015].
However, inference from population models is typically limited to small spatial extents because the data required is difficult to collect across wide species ranges.
Indeed, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [see, e.g., @Salguero-Gomez2015].
Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
Thus, the real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011, Tredennick and Adler in review].

@@

Large-scale, spatially explicit population models based on RS data could offer a valuable new way to investigate the effects of large-scale environmental changes--especially climate change--playing out at landscape and regional scales.  
Currently most assesments of how plant and animal populations will respond to climate change rely on Species distribution models (SDMs).  
SDMs rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009] and they are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014]. 
However, the short-term and small-scale population dynamics that actually drive the large-scale distributions of species are not represented in most SDMs. 
Since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [add EhrlenANDMorris]. 
Furthermore, because they lack short-term dynamics they usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached within our lifetimes. Large-scale dynamic population models could overcome both of these limitations.  
They would produce spatially expicit estimates of species abundance within the species range [Ehrlen and Morris] and would provide testable predictions of how populations should respond to short-term climate perturbations.  
These short-term predictions would give ecological forecasters the opportunity to repeatedly validate and refine their models (cite a prediction paper).

@@

Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data, we demonstrate how large-scale plant population models could be used to predict how climate change will affect sagebrush and other species in the future. 
Our approach relies on a process model motivated by Gompertz density-dependent population growth and a cutting-edge remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012]. 
We rigorously account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the largest potential.
In those cases, our approach is particularly well-suited for combining the information of population models (e.g., population status) and the spatial extent of species distribution models.

@@

Conclusion
----------
We have introduced a new approach to fitting and simulating population models at large spatial extents with plant population data derived from state of the art remote sensing.
As more species-level remote sensing datasets become available and computing power increases this approach will be applicable to a wider number of species and even larger spatial extents.
Future modeling could include the effects of non-climate drivers--including the effects of species interactions and disturbance [@Clark2014].  
For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford2009].  
Fortunately, our spatio-temporal modeling framework could easily be extended to model additional species and dynamic processes as the data become availabe. 
The approach we have developed here fills an important gap in spatial scales between species distribution models and local-scale demographic population models.

>>>>

Abstract
--------
New remote sensing technology is producing ecological data at unprecedented spatial scales. 
These massive data offer an opportunity to study ecological processes at the landscape and regional scales that have traditionally been observed only at the level of small study plots. 
However, harnessing the power of new spatial data requires statistical models that can both rigorously account for spatial autocorrelation and can also be fit within a reasonable amount of computation time. 
Here we demonstrate how recent advances in the modeling of animal populations can be applied to remotely-sensed, species-specific estimates of plant cover at the landscape scale.
Using modern, computationally efficient model reparameterizations we fit a spatiotemporal population model to a 28 year time series of sagebrush (*Artemisia* spp.) percent cover over a $2.5\times5$ km landscape in southwestern Wyoming with rigorous accounting of spatial autocorrelation.
We include interannual variation in precipitation and temperature as covariates in the model to investigtate how climate affects the cover of this species.
We then use the model to forecast the future abudance of sagebrush at the landscape scale under projected climate change. 
Our analysis provides spatially explicit estimates of change in a plant population that have, until now, been impossible to produce at this scale. 
Critically our large-scale and long-term predictions are rooted in small-scale and short-term population dynamics and so serve as a useful alternative to the predictions offered by most species distribution models that do not include population dynamics at all. 
We expect the approach we develop here will become an indispensable method for using new remote sensing data to model how populations respond to environmental changes, especially climate change, that play out at spatial scales far greater than the average study plot.

@@

Introduction
-------------
Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology [@Clark2001; @Petchey2015].
Population models are ideally suited for meeting such a challenge because they allow inference on population abundance and dynamics (year-to-year changes in abundance) in relation to climate drivers [@Hare2010; @Adler2012; @Ross2015].
However, inference from population models is typically limited to small spatial extents because the data required is difficult to collect across wide species ranges.
Indeed, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale [see, e.g., @Salguero-Gomez2015].
Local-scale demographic data make building population projection models an easy task [@Ellner2006; @Rees2009; @Adler2012], but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions [@Freckleton2011; @Queenborough2011].
Thus, the real challenge is not to simply make population forecasts, but to do so at spatial scales relevant to policy and management decisions [@Queenborough2011; @Tredennick2015].

@@

Large-scale, spatially explicit population models based on RS data could offer a valuable new way to investigate the effects of large-scale environmental changes--especially climate change--playing out at landscape and regional scales.
Currently most assesments of how plant and animal populations will respond to climate change rely on species distribution models (SDMs).
SDMs rely on static associations between contemporary climate and a species' distribution or, more rarely, abundance to project future distribution or abundance [@Elith2009] and they are easily applied at landscape to continental scales [e.g., @Maiorano2013; @Clark2014]. 
However, the short-term and small-scale population dynamics that actually drive the large-scale distributions of species are not represented in most SDMs. 
Since SDMs typically rely on occurence data, their projections of habitat suitability or probability of occurence provide little information on the future states of populations in the core of their range -- areas where a species exists now and is expected to persist in the future [@Ehrlen2015]. 
Furthermore, because they lack short-term dynamics they usually cannot produce any estimate of the rate at which local populations will increase or decrease in the near-term and instead project a future equilibrium species distribution that may or may not ever be reached within our lifetimes.
Large-scale dynamic population models could overcome both of these limitations.
They would produce spatially explicit estimates of species abundance within the species range [@Ehrlen2015] and would provide testable predictions of how populations should respond to short-term climate perturbations.
These short-term predictions would give ecological forecasters the opportunity to repeatedly validate and refine their models [@Luo2011a].

@@

Building on recent technological advances in spatial statistics and anticipating ever-increasing availability of RS data [@He2015], we demonstrate how large-scale plant population models could be used to predict how climate change will affect sagebrush and other species in the future. 
Our approach relies on a process model motivated by Gompertz density-dependent population growth and a cutting-edge remotely-sensed time series of sagebrush cover from Wyoming [@Homer2012; @Homer2015]. 
We rigorously account for spatial autocorrelation with dimension reduction techniques borrowed from the animal modeling literature [@Conn2015] and produce spatially-explicit estimates of sagebrush percent cover.
The modeling framework we propose can be applied to any spatially-explicit time series of plant cover or density, but its application to remotely-sensed data products offers the largest potential.
In those cases, our approach is particularly well-suited for combining the information of population models (e.g., population status) and the spatial extent of species distribution models.

@@

Conclusion
----------
We have introduced a new approach to fitting and simulating population models at large spatial extents with plant population data derived from state of the art remote sensing.
As more species-level remote sensing datasets become available and computing power increases this approach will be applicable to a wider number of species and even larger spatial extents.
Future modeling could include the effects of non-climate drivers--including the effects of species interactions and disturbance [@Clark2014].
For sagebrush, including fire and competition with non-native annual grasses in the model may be especially important for a complete assesment of the effects of climate change [@Bradford2006].
Fortunately, our spatio-temporal modeling framework could easily be extended to model additional species and dynamic processes as the data become availabe.
The approach we have developed here fills an important gap in spatial scales between species distribution models and local-scale demographic population models.
